---
title: "Tutorial: Simulate Tree Rings"
author: "Daniel J. Hocking & Laura G. Smith"
date: "`r Sys.Date()`"
output:
      rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Tutorial: Simulate Tree Rings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, fig.show='hold', echo=FALSE, eval=FALSE}
# The figure sizes have been customised so that you can easily put two images side-by-side.
plot(1:10)
plot(10:1)

# Yyou can use the chunk option `fig.cap = "Your figure caption."`
```


```{r, echo=FALSE, results='asis', eval=FALSE}
# tables, e.g. using `knitr::kable()`.
knitr::kable(head(mtcars, 10))
```

```{r, warning=FALSE, message=FALSE, eval=FALSE}
library(devtools)
install_github(repo = "lgsmith295/simDendro")
```


## Simulate Negative Exponential Growth

$$
rwl = biological \times climate \times error
$$

### Modified Negative Exponential

$$
rwl_{it} = ae^{-bt + k} e^{\beta \times climate}e^{error}
$$

$$
log(rwl_{it}) = \alpha - bt + \beta \times climate + error
$$

where $\alpha = log(a) + k$

$$
log(rwl_{it}) = \alpha_{0it} + \alpha_{1it} age_{it} + \beta X_{it} + \epsilon_{it}
$$


```{r, fig.width=8, fig.height=4}
library(simDendro)
years <- 1:50

# Growth
growth <- negexp_growth(length(years), 1, 0.03, 0.001)
plot(growth, type = "b")

# climate random, uncorrelated around mean
mu_climate <- 12
climate <- rnorm(length(years), mu_climate, 3)
plot(climate, type = "l")

# Climate effect on growth
beta_mu <- 0.8 # real scale slope - 1 degree C increase in T results in this change in mm tree ring growth
# log_beta <- log(beta_mu)
beta <- rnorm(length(years), beta_mu, 0.05)
eta <- beta * climate
plot(eta, type = "l")
plot(climate, eta)

# IID error
eps <- rnorm(length(years), -2.5, 0.5)
error <- exp(eps)
summary(error)
plot(error, type = "l")

# Put growth model together
rwl <- growth * eta * error
plot(rwl, type = "b")

```

## climate with linear trend and random noise

```{r, fig.width=8, fig.height=4}
library(simDendro)
years <- 1:50

# Growth
growth <- negexp_growth(length(years), 1, 0.03, 0.1)
plot(growth, type = "b")

# climate
mu_climate <- 10 + 0.1 * years
climate <- rnorm(length(years), mu_climate, 3)
plot(climate, type = "l")

# Climate effect on growth
beta_mu <- 0.8 # real scale slope - 1 degree C increase in T results in this change in mm tree ring growth
# log_beta <- log(beta_mu)
beta <- rnorm(length(years), beta_mu, 0.05)
eta <- beta * climate
plot(eta, type = "l")
plot(climate, eta)

# IID error
eps <- rnorm(length(years), -2.5, 0.5)
error <- exp(eps)
summary(error)
plot(error, type = "l")

# Put growth model together
rwl <- growth * eta * error
plot(rwl, type = "b")

```

## Allow climate to vary over time

```{r}
years <- 1:500

# Growth
growth <- negexp_growth(length(years), 1, 0.03, 0.1)
plot(growth, type = "b")

# Climate: autoregressive model around set mean
mu_climate <- 12
yt <- arima.sim(n=500, list(order=c(1,0,0),ar=0.9), sd = 0.5) + mu_climate  # adding intercept and trend to the autoregressive 
plot(yt)
acf(yt)

### Climate: autoregressive model with linear trend
mu_climate <- 12
yt <- arima.sim(n=500, list(order=c(1,0,0),ar=0.9), sd = 0.5) + mu_climate + years*0.01  # adding intercept and trend to the autoregressive 
plot(yt)
acf(yt)


### Climate: basis splines
mu_climate <- 12

library(mgcv)
x <- seq(0,500,length=501)
knots1 <- seq(0, 500, by = 100)
sm <- smoothCon(s(x,bs="bs"), data.frame(x), knots = data.frame(knots1))[[1]] # addition of knots function not working and not throwing warning or error
matplot(1:501, sm$X, type="l",xlab="Time",ylab="Basis function, Bj(X)",cex.lab=1.5,cex.axis=1.5,lwd=2)

set.seed(416923)
beta  <- rnorm(ncol(sm$X), 0, 3)
g <- (sm$X %*% beta) + mu_climate
plot(g, type = "l")

# add random noise around trend
climate <- g + rnorm(length(g), 0, 2)
plot(climate, type = "l")
acf(climate)

# Climate effect on growth
beta_mu <- 0.8 # real scale slope - 1 degree C increase in T results in this change in mm tree ring growth
# log_beta <- log(beta_mu)
beta <- rnorm(length(years), beta_mu, 0.05)
eta <- beta * climate
plot(eta, type = "l")
plot(climate, eta)

# IID error
eps <- rnorm(length(years), -2.5, 0.5)
error <- exp(eps)
summary(error)
plot(error, type = "l")

# Put growth model together
rwl <- growth * eta * error
plot(rwl, type = "b")


```

## Create series for multiple trees with the same climate

```{r}
##### NegExp Growth - Mean Climate #####
years <- 1:500

# Growth
growth <- negexp_growth(length(years), 1, 0.03, 0.001)
plot(growth, type = "b")

# climate random, uncorrelated around mean
mu_climate <- 12
climate <- rnorm(length(years), mu_climate, 3)
plot(climate, type = "l")

# Climate effect on growth
beta_mu <- 0.8 # real scale slope - 1 degree C increase in T results in this change in mm tree ring growth
# log_beta <- log(beta_mu)
beta <- rnorm(length(years), beta_mu, 0.05)
eta <- beta * climate
plot(eta, type = "l")
plot(climate, eta)

# IID error
eps <- rnorm(length(years), -2.5, 0.5)
error <- exp(eps)
summary(error)
plot(error, type = "l")

# Put growth model together
rwl <- growth * eta * error
plot(rwl, type = "b")


# loop over multiple trees
M <- 100 # number of trees
Tea <- 500
years <- 1500 + 1:Tea

# Growth
tgrowth <- matrix(NA, nrow = Tea, ncol = M)
# assume each tree lives 200 years for simplicity and germinated in a random year between 1350 and 1950 - assumes can't get to pith in older trees
fy <- floor(runif(M, 1351, 1951))
k <- rlnorm(M, log(0.2), 0.2) # tree biological growth curves vary
b <- rlnorm(M, log(0.04), 0.1) # tree biological growth curves vary
a <- rlnorm(M, log(1), 0.2) # tree biological growth curves vary - maybe should be multivariate normal between the parameters
hist(a)
hist(b)
hist(k)
sd(b)

for(i in 1:M) {
  bio <- negexp_growth(200, a[i], b[i], k[i])
  yrs <- fy[i]:(fy[i] + 199)
  df_tmp <- data.frame(tree = i, year = yrs, bio)
  if(i == 1) {
    df <- df_tmp
  } else {
    df <- bind_rows(df, df_tmp)
  }
}

str(df)

# climate random, uncorrelated around mean
mu_climate <- 12
climate <- rnorm(Tea, mu_climate, 3)
plot(climate, type = "l")

# Climate effect on growth
mu_beta <- 0.8 # real scale slope - 1 degree C increase in T results in this change in mm tree ring growth
# log_beta <- log(beta_mu)
beta <- rnorm(Tea, mu_beta, 0.05) # this should probably vary by tree but I guess that's going into other variance components
eta <- beta * climate
plot(eta, type = "l")
plot(climate, eta)
eta <- data.frame(year = years, eta, climate)

# IID error
eps <- rnorm(Tea, -2.5, 0.5)
error <- exp(eps)
summary(error)
plot(error, type = "l")

# Random IID tree variance among years
sigma_eps_i <- rlnorm(M, log(0.5), 0.05)
hist(sigma_eps_i)
eps_ij <- data.frame(expand.grid(tree = 1:M, year = years, eps = NA_real_)) %>%
  dplyr::arrange(tree, year)
for(i in 1:M) {
  eps_ij[which(eps_ij$tree == i), ]$eps <- rlnorm(Tea, log(0.08), sigma_eps_i[i])
}
hist(eps_ij$eps)
# tree_sd <- data.frame(tree = 1:M, eps_i = eps_i)

# Put model together
rwl_long <- df %>%
    group_by(tree) %>%
    left_join(eta) %>%
    left_join(eps_ij) %>%
    mutate(rwl = bio * eta * eps,
           rwi = rwl - bio,
           climate_std = (climate - mean(climate, na.rm = TRUE)) / sd(climate, na.rm = TRUE)) %>%
    dplyr::filter(year %in% years)

ggplot(rwl_long, aes(year, rwl)) + geom_line(alpha = 0.3, color = "blue") + theme(legend.position = "none") + theme_bw()

ggplot(rwl_long, aes(year, rwi)) + geom_point(alpha = 0.1, color = "turquoise") + theme(legend.position = "none") + theme_bw()

ggplot(rwl_long, aes(year, rwi)) + 
  geom_point(alpha = 0.1, color = "turquoise", aes(group = tree)) + 
  geom_smooth() + 
  geom_line(aes(year, climate_std), alpha = 0.1) + 
  geom_smooth(aes(year, climate_std), color = "red", alpha = 0.5) + 
  theme(legend.position = "none") + 
  theme_bw()

# just show 1st 10 trees
rwl_small <- rwl_long %>%
  dplyr::filter(tree %in% 11:20)

ggplot(rwl_small, aes(year, rwl, group = tree, color = as.factor(tree))) + geom_line(alpha = 0.3) + theme_bw() + theme(legend.position = "none")


##### Allow climate to vary over time #####

# Growth
growth <- negexp_growth(length(years), 1, 0.03, 0.1)
plot(growth, type = "b")

# Climate: autoregressive model around set mean
mu_climate <- 12
yt <- arima.sim(n=500, list(order=c(1,0,0),ar=0.9), sd = 0.5) + mu_climate  # adding intercept and trend to the autoregressive 
plot(yt)
acf(yt)

### Climate: autoregressive model with linear trend
mu_climate <- 12
yt <- arima.sim(n=500, list(order=c(1,0,0),ar=0.9), sd = 0.5) + mu_climate + years*0.01  # adding intercept and trend to the autoregressive 
plot(yt)
acf(yt)


### Climate: basis splines
mu_climate <- 12

library(mgcv)
x <- seq(0,500,length=501)
knots1 <- seq(0, 500, by = 100)
sm <- smoothCon(s(x,bs="bs"), data.frame(x), knots = data.frame(knots1))[[1]] # addition of knots function not working and not throwing warning or error
matplot(1:501, sm$X, type="l",xlab="Time",ylab="Basis function, Bj(X)",cex.lab=1.5,cex.axis=1.5,lwd=2)

set.seed(416923)
beta  <- rnorm(ncol(sm$X), 0, 3)
g <- (sm$X %*% beta) + mu_climate
plot(g, type = "l")

# add random noise around trend
climate <- g + rnorm(length(g), 0, 2)
plot(climate, type = "l")
acf(climate)

# loop over multiple trees
M <- 100 # number of trees
Tea <- 500
years <- 1500 + 1:Tea

# Growth

# assume each tree lives 200 years for simplicity and germinated in a random year between 1350 and 1950 - assumes can't get to pith in older trees
fy <- floor(runif(M, 1351, 1951))
k <- rlnorm(M, log(0.2), 0.2) # tree biological growth curves vary
b <- rlnorm(M, log(0.04), 0.1) # tree biological growth curves vary
a <- rlnorm(M, log(1), 0.2) # tree biological growth curves vary - maybe should be multivariate normal between the parameters
hist(a)
hist(b)
hist(k)
sd(b)

for(i in 1:M) {
  bio <- negexp_growth(200, a[i], b[i], k[i])
  yrs <- fy[i]:(fy[i] + 199)
  df_tmp <- data.frame(tree = i, year = yrs, bio)
  if(i == 1) {
    df <- df_tmp
  } else {
    df <- bind_rows(df, df_tmp)
  }
}

str(df)

# climate spline
mu_climate <- 12

library(mgcv)
x <- years
knots1 <- seq(0, max(years), by = 100)
sm <- smoothCon(s(x,bs="bs"), data.frame(x), knots = data.frame(knots1))[[1]] # addition of knots function not working and not throwing warning or error

set.seed(416923)
beta  <- rnorm(ncol(sm$X), 0, 3)
g <- (sm$X %*% beta) + mu_climate
# plot(g, type = "l")

# add random noise around trend
climate <- g + rnorm(length(g), 0, 0.5)
# plot(climate, type = "l")
# acf(climate)

# Climate effect on growth
mu_beta <- 0.8 # real scale slope - 1 degree C increase in T results in this change in mm tree ring growth
# log_beta <- log(beta_mu)
beta <- rnorm(Tea, mu_beta, 0.05) # this should probably vary by tree but I guess that's going into other variance components
eta <- beta * climate
# plot(eta, type = "l")
# plot(climate, eta)
eta <- data.frame(year = years, eta, climate)

# IID error
eps <- rnorm(Tea, -2.5, 0.5)
error <- exp(eps)
# summary(error)
# plot(error, type = "l")

# Random IID tree variance among years
sigma_eps_i <- rlnorm(M, log(0.5), 0.05)
hist(sigma_eps_i)
eps_ij <- data.frame(expand.grid(tree = 1:M, year = years, eps = NA_real_)) %>%
  dplyr::arrange(tree, year)
for(i in 1:M) {
  eps_ij[which(eps_ij$tree == i), ]$eps <- rlnorm(Tea, log(0.08), sigma_eps_i[i])
}
# hist(eps_ij$eps)
# tree_sd <- data.frame(tree = 1:M, eps_i = eps_i)

# Put model together
rwl_long <- df %>%
  group_by(tree) %>%
  left_join(eta) %>%
  left_join(eps_ij) %>%
  mutate(rwl = bio * eta * eps,
         rwi = rwl - bio,
         rwi_div = rwl / bio,
         climate_std = (climate - mean(climate, na.rm = TRUE)) / sd(climate, na.rm = TRUE)) %>%
  dplyr::filter(year %in% years)

ggplot(rwl_long, aes(year, rwl)) + geom_line(alpha = 0.3, color = "blue") + theme(legend.position = "none") + theme_bw()

ggplot(rwl_long, aes(year, rwi)) + geom_point(alpha = 0.1, color = "turquoise") + theme(legend.position = "none") + theme_bw()

ggplot(rwl_long, aes(year, rwi)) + 
  geom_point(alpha = 0.1, color = "turquoise", aes(group = tree)) + 
  geom_smooth() + 
  geom_line(aes(year, climate_std), alpha = 0.1) + 
  geom_smooth(aes(year, climate_std), color = "red", alpha = 0.5) + 
  theme(legend.position = "none") + 
  theme_bw()

ggplot(rwl_long, aes(year, rwi_div)) + 
  geom_point(alpha = 0.1, color = "turquoise", aes(group = tree)) + 
  geom_smooth() + 
  geom_line(aes(year, climate_std), alpha = 0.1) + 
  geom_smooth(aes(year, climate_std), color = "red", alpha = 0.5) + 
  theme(legend.position = "none") + 
  theme_bw()

# just show 1st 10 trees
rwl_small <- rwl_long %>%
  dplyr::filter(tree %in% 11:20)

ggplot(rwl_small, aes(year, rwl, group = tree, color = as.factor(tree))) + geom_line(alpha = 0.3) + theme_bw() + theme(legend.position = "none")

```


## Partial Samples (not near pith) - maybe move this to own repository using the package and separate from the vignette

Simulate all tree growth using a biological growth model but then cut off the samples to more recent years due to things like heart rot, alternative growth forms, and missing the pith. See how different models assuming different growth (e.g. negex, spline, linear, mean) recover the data and reconstruct climate.

